\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-19 paper.qmd
Updated upper limit of b1
Updated Figure 15

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \frac{n}{2}$ are allowed. @fig-param-scurve (b) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_15_min_dist_0.99.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds") 
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds") 
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2,
                        error_pbmc_phate,
                        error_pbmc_trimap,
                        error_pbmc_pacmap)
```

```{r}
#| label: tsne-pbmc
# read the png file from device 
tsne_5_img <- readPNG("figures/pbmc3k/tsne_5.png", native = TRUE) 
phate_5_img <- readPNG("figures/pbmc3k/phate_5.png", native = TRUE) 
umap_img <- readPNG("figures/pbmc3k/author_view.png", native = TRUE) 
tsne_30_img <- readPNG("figures/pbmc3k/tsne_30.png", native = TRUE)
umap_15_img <- readPNG("figures/pbmc3k/umap_15.png", native = TRUE) 
pacmap_img <- readPNG("figures/pbmc3k/pacmap.png", native = TRUE) 
trimap_img <- readPNG("figures/pbmc3k/trimap.png", native = TRUE)
umap_5_img <- readPNG("figures/pbmc3k/umap_5.png", native = TRUE)

## Add arrows
arrow_df_pbmc <- tibble(x = c(94, 280, 101, 150, 218, 250, 50),
             xend = c(96, 295, 96, 180, 250, 290, 25),
             y = c(10.06, 9.69, 9.88, 9.855, 9.82, 9.78, 10),
             yend = c(10.14, 9.8, 9.8, 10, 10.3, 10, 9.8))

error_plot_pbmc <- ggplot(error_pbmc, 
                              aes(x = b_non_empty, 
                                  y = log(Error), 
                                  colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 135, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")+
    geom_segment(data = arrow_df_pbmc, aes(
      x = x,  
      y = y,  
      xend = xend, 
      yend = yend), 
      arrow = arrow(length = unit(0.01, "npc"),
        ends = "first"), 
        color = "black",
      linewidth = 0.1) +
  theme(#aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        legend.position = "bottom",
        legend.text = element_text(size = 7),
        legend.key.height = unit(0.3, 'cm'),
        legend.key.width = unit(0.3, 'cm'),
        legend.key.size = unit(3, 'cm')) +
    inset_element(p = tsne_5_img, 
                  left = 1.8, 
                  bottom = 0.08, 
                  right = 0, 
                  top = 0.25) +                   
    inset_element(p = phate_5_img, 
                  left = 0.6, 
                  bottom = 0.4, 
                  right = 0, 
                  top = 0.55) +                   
    inset_element(p = umap_img, 
                  left = 1.2, 
                  bottom = 0.3, 
                  right = -0.1, 
                  top = 0.5) +
    inset_element(p = umap_15_img, 
                  left = 1.5, 
                  bottom = 0.5, 
                  right = 0, 
                  top = 0.7) +
    inset_element(p = tsne_30_img, 
                  left = 0.6, 
                  bottom = 0.01, 
                  right = 0, 
                  top = 0.18) +
    inset_element(p = pacmap_img, 
                  left = 1.7, 
                  bottom = 0.3, 
                  right = 0, 
                  top = 0.5) +
    inset_element(p = umap_5_img, 
                  left = 0.4, 
                  bottom = 0.05, 
                  right = -0.2, 
                  top = 0.22)

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $136$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 70%

>>>>

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{\frac{n}{2}}$ are allowed. @fig-param-scurve (b) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_15_min_dist_0.99.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
# error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds") 
# error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds") 
# error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2)

error_pbmc <- error_pbmc |>
  filter(bin1 >= 5)
```

```{r}
#| label: tsne-pbmc
# read the png file from device

pbmc_layouts <- readPNG("figures/pbmc3k/pbmc_nldr_layouts.png", native = TRUE)

error_plot_pbmc <- ggplot(error_pbmc, 
                          aes(x = b_non_empty, 
                              y = Error/1000, 
                              colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 136, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#ff7f00', '#984ea3', 
                              '#999999', '#a65628', '#f781bf')) +
  ylab("absolute error (in thousands)") +
  xlab("number of non-empty bins") + 
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        axis.ticks.x = element_line(linewidth = 0.5),
        axis.ticks.y = element_line(linewidth = 0.5),
        legend.position = "none") +
  inset_element(p = pbmc_layouts, 
                left = 1.55, 
                bottom = 0.35, 
                right = 0, 
                top = 1)

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to PBMC3k dataset with diffferent (hyper-)parameter choices. What is the best (hyper-)parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $136$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 70%

### 2024-08-19 paper.tex
Updated error plot
Updated upper limit of b1
Updated Figure 15

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-scurve-1.pdf}

@@

\caption{\label{fig-p-d-error-in-2d-scurve}}

@@

The default number of bins \(b=b_1\times b_2\) is computed based on the
sample size, by setting \(b_1=n^{1/3}\), consistent with the
Diaconis-Freedman rule \citep{freedman1981}. The value of \(b_2\) is
determined analytically by \(b_1, q, r_2\). Values of \(b_1\) between
\(2\) and \(b_1 = \frac{n}{2}\) are allowed.
Figure~\ref{fig-param-scurve} (b) shows the effect of different choices
of \(b_1\) on the MSE of the fitted model.

@@

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to \(136\), the
lowest error occurred with the parameters perplexity: \(30\).}

>>>>

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-scurve-1.pdf}

@@

\caption{\label{fig-p-d-error-in-2d-scurve}error}

@@

The default number of bins \(b=b_1\times b_2\) is computed based on the
sample size, by setting \(b_1=n^{1/3}\), consistent with the
Diaconis-Freedman rule \citep{freedman1981}. The value of \(b_2\) is
determined analytically by \(b_1, q, r_2\). Values of \(b_1\) between
\(2\) and \(b_1 = \sqrt{\frac{n}{2}}\) are allowed.
Figure~\ref{fig-param-scurve} (b) shows the effect of different choices
of \(b_1\) on the MSE of the fitted model.

@@

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to PBMC3k dataset with diffferent (hyper-)parameter choices.
What is the best (hyper-)parameter choice to create the model? The
residual plot have a steep slope at the beginning, indicating that a
smaller number of non-empty bins causes a larger amount of error. Then,
the slope gradually declines or level off, indicating that a higher
number of non-empty bins generates a smaller error. Using the elbow
method, it was observed that when the number of non-empty bins is set to
\(136\), the lowest error occurred with the parameters perplexity:
\(30\).}
